---
title: 'Business Statistics End of Term Assessment IB94X0 2025-2026 #1'
author: '5754848'
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(tidyverse)
library(gridExtra)
library(broom) 
library(kableExtra) 
library(ggplot2)
library(stringr) 
library(ggeffects)
library(modelsummary)
library(plotly)
```


# Academic Integrity Statement

I confirm that this submission is my own work. I used generative AI (Google Gemini AI Pro) to help draft and refine code snippets and to proofread explanatory text; any AI outputs were edited and integrated by me to ensure accuracy and originality.I have not plagiarized the work of others; . Because the assignment permits AI, I used it in an appropriate and ethical manner, and I retain intellectual ownership of the final submission.


# Disclaimer on Use of Generative Artificial Intelligence

This assignment is the result of the author’s independent work, with only minimal and clearly delineated use of generative artificial intelligence tools (Google Gemini AI Pro). AI assistance was employed solely to explore the implementation of selected R packages that were not covered in the workshops or online lectures. This support was limited to enhancing the visual presentation of plots, improving aesthetic quality, and ensuring that the report adhered to the technical report guidelines specified in the assignment brief.

All analytical reasoning, methodological choices, coding logic, and interpretation of results were developed independently by the author in accordance with the academic expectations of the module. Any AI‑suggested elements were critically evaluated, fully understood, and appropriately adapted before inclusion. The final workflow and outputs reflect the techniques, principles, and standards taught throughout the course.

AI support was also used in a restricted capacity to refine grammar, clarity, and overall presentation within code comments and report text. The AI did not generate analytical results, perform computations, or contribute to the interpretation of findings. Its role was strictly supplementary, confined to formatting alignment and professional presentation.

---

# Question 1

This report provides a comprehensive statistical analysis of weekly household food expenditure in England, investigating how various socioeconomic and demographic factors influence domestic spending patterns. Using a dataset of individual household records. This study examines the relationship between food costs and predictors including household size, housing expenses, vehicle ownership, income category, and the sex of the respondent. Beyond general trends, a specific analytical focus is placed on the impact of alcohol expenditure on the food budget and whether this relationship is moderated by a household’s existing debt obligations. By applying multiple regression modeling and interaction analysis, this report aims to identify the primary drivers of food expenditure while determining if the financial pressure of debt alters the spending behavior associated with alcohol consumption.

### Data Dictionary

The variables from the data provided are described in the table below

| Variable | Description |
|------------------------------------|------------------------------------|
| Kids | No of Kids in a Household |
| HouseholdSize | Size of the Household. |
| Sex | Sex of the survey respondent |
| Vehicles | No of Vehicles in a Household |
| Food | The average amount spent on food per week |
| Housing | The average amount spent on Housing per week |
| Alcohol | The average amount spent on Alcohol per week |
| Debt | The average amount spent on Debt per week |
| income_category | Income Category of the Household |

First, the "HouseholdExpenditure.csv" dataset is imported into RStudio and assigned to a tibble named household_main, which will serve as the primary object for subsequent data exploration and analysis.

```{r}
#Import the HouseholdExpenditure dataset into R as a tibble named 'household_main'
household_main <- read_csv("HouseholdExpenditure.csv")
```

### Exploratory Data Analysis

We note that the dataset contains two nominal categorical variables, Sex and income_category. To further examine the structure and data types within the dataset, we employ the str() function to obtain a concise, technical overview of all variables.

```{r}
# Inspect the structure of 'household_main' to view data types, 
# number of observations, and a preview of each variable
str(household_main)
```
We identify three ordinal categorical variables in the dataset—kids, householdsize, and vehicles—alongside four continuous numerical variables: Food, Alcohol, Debt, and Housing.Then, We create a separate dataset, expenditure_vars, containing only the expenditure‑related variables to simplify and streamline the analysis.

```{r}
# Select only expenditure variables
expenditure_vars <- household_main %>%
  select(Food, Housing, Alcohol, Debt)
```

We generate bar plots for the ordinal categorical variables (kids, householdsize, and vehicles) to examine how observations are distributed across the categories within each variable.

```{r}
# --- Distribution of Categorical Predictors ---

household_main %>%
  # Reshape data to long format to allow simultaneous plotting via facets
  pivot_longer(
    cols = c(Kids, HouseholdSize, Vehicles),
    names_to = "variable",
    values_to = "value"
  ) %>%
  # Treat numeric categories as factors for discrete bar representation
  ggplot(aes(x = as.factor(value))) +
  # Create bar charts to visualize frequency counts
  geom_bar(fill = "#2C7FB8", alpha = 0.85) +
  # Split plot by variable; scales="free_x" handles different ranges (e.g., 1-5 vs 0-2)
  facet_wrap(~ variable, scales = "free_x") +
  # Professional labeling for public-report readiness [cite: 17]
  labs(
    title = "Distribution of Categorical Household Variables",
    subtitle = "Initial check for data frequency and potential outliers",
    x = "Category Level",
    y = "Household Count"
  ) +
  theme_minimal()
```


The bar chart titled “Distribution of Categorical Household Variables” shows that most households fall within the lower categories of the ordinal variables HouseholdSize, Kids, and Vehicles. Household sizes are predominantly between 1 and 3 members, with larger households being relatively rare. Similarly, most families report zero or one child, and the number declines steadily as the number of children increases. Vehicle ownership peaks at one vehicle, followed by zero and two, with higher counts being uncommon. These patterns highlight the concentration of observations in lower categories and indicate that sparse upper‑level categories may require careful handling or consolidation in subsequent modelling.

We construct histograms for the expenditure variables (Alcohol, Food, Debt, and Housing) to examine and better understand the underlying distribution of each variable.

```{r}
# --- Expenditure Distribution ---

expenditure_vars %>%
  # Reshape data for efficient plotting using the tidyverse 'pivot_longer' function
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  # Initialize ggplot for visual inspection of Food, Housing, Alcohol, and Debt distributions
  ggplot(aes(x = value)) +
  # Use 60 bins for better visualization and for outlier detection
  geom_histogram(fill = "#2C7FB8", color = "white", bins = 60, alpha = 0.85) +
  # Facet by variable with free scales to accommodate different expenditure ranges
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  # Professionally format titles and labels suitable for external publication (Source 17)
  labs(
    title = "Distribution of Household Expenditure Variables",
    subtitle = "Faceted histograms for Food, Housing, Alcohol, and Debt",
    x = "Weekly Expenditure (£)",
    y = "Frequency (Household Count)"
  ) +
  # Apply theme enhancements for clarity and professional presentation
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#08306B"),
    plot.subtitle = element_text(size = 12, color = "#08519C"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85")
  )
```


The histograms show distinct patterns across the expenditure variables. Alcohol and Food display relatively symmetric distributions with clear central peaks, indicating consistent spending. Debt is right-skewed, with a spike at £0 and a long tail, suggesting most households carry low or no debt, while a few have high debt. Housing also shows right-skewness and wider spread, reflecting greater variability in housing costs. These patterns highlight the need to consider potential transformations or modelling adjustments for skewed variables.

Although the histograms offer a helpful overview of the distributions for both the expenditure variables and the ordinal categorical variables (kids, householdsize, and vehicles), they only provide a general indication of potential outliers. To identify specific outlier values more precisely, we construct boxplots for the expenditure_vars tibble as well as for the ordinal categorical variables.

```{r}
# --- Outlier Detection and Distribution ---

# 1. Boxplots for Continuous Expenditure Variables

expenditure_vars %>%
  # Reshape data for faceted plotting of Food, Housing, Alcohol, and Debt
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  # Use professional styling for outliers to aid visual inspection
  geom_boxplot(
    fill = "#74A9CF",
    color = "#045A8D",
    width = 0.6,
    outlier.color = "#023858",
    outlier.size = 1.8,
    alpha = 0.9) + 
  # Separate plots with independent y-axes to accommodate different frequency ranges
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Boxplots of Expenditure Variables",
    subtitle = "Visual inspection of distributions and potential outliers across numeric fields",
    x = "",
    y = "Value (£)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#08306B"),
    plot.subtitle = element_text(size = 12, color = "#08519C"),
    strip.text = element_text(face = "bold", size = 12, color = "white"),
    strip.background = element_rect(fill = "#045A8D", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    panel.border = element_rect(fill = NA, color = "grey70", size = 0.5)
  )



# 2. Boxplots for Discrete Household Characteristics

household_main %>%
  pivot_longer(
    cols = c(Kids, HouseholdSize, Vehicles),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(
    fill = "#74A9CF",
    color = "#045A8D",
    width = 0.6,
    outlier.color = "#023858",
    outlier.size = 1.8,
    alpha = 0.9
  ) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Boxplots of Household Characteristics",
    subtitle = "Distribution check for Kids, Household Size, and Vehicles",
    x = "",
    y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#08306B"),
    plot.subtitle = element_text(size = 12, color = "#08519C"),
    strip.text = element_text(face = "bold", size = 12, color = "white"),
    strip.background = element_rect(fill = "#045A8D", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    panel.border = element_rect(fill = NA, color = "grey70", size = 0.5)
  )
```


The boxplots show clear outliers across both expenditure and ordinal variables. Alcohol and Food have a few extreme values on both ends, while Debt displays the largest number of outliers, especially at very high amounts. Housing also shows several unusually low and high values beyond the typical range. For the ordinal variables, Kids, HouseholdSize, and Vehicles all have outliers extending well above their common categories, indicating a small number of unusually large households, large families, or high vehicle ownership. These outliers may require careful consideration in later modelling.

Next, we examine the number of missing values (NAs) in each variable to determine which observations should be removed, as these incomplete entries are not required for subsequent analysis.

```{r}
#calculates the total count of missing (NA) values for each variable in the dataset.
colSums(is.na(household_main))
```

We remove all observations containing missing values from household_main and create a cleaned tibble, householdmain_clean, which contains only complete cases for subsequent analysis.


```{r}
#Removes all the NA values from all the columns
(householdmain_clean <- household_main %>% drop_na())
```


We also perform a quick check for any misspellings or inconsistencies in the categorical variables Sex and income_category to ensure that all entries are clean and uniformly recorded.

```{r}
# Check unique entries for gender to confirm any misspelled entry/duplicate entry
unique(householdmain_clean$Sex)

# Check unique entries for income to ensure the 'high/medium/low' structure
unique(householdmain_clean$income_category)
```

Manual encoding was chosen instead of relying on automatic encoding to maintain full control over how each category is translated into numeric form. This ensures that the encoded variables follow a structure appropriate for regression modelling, avoid accidental creation of redundant dummy variables, and remain consistent with the analytical framework used throughout the study. By defining sex_num and income_num manually, we ensure clean, interpretable, and model‑ready variables that support a more reliable multiple regression analysis.

```{r}
householdmain_clean <- householdmain_clean %>%
  mutate(
    # Create a binary numeric variable for Sex (Male = 1, Female = 0).
    # str_trim and str_to_lower are used to handle potential whitespace or 
    # casing inconsistencies, ensuring data integrity.
    sex_num = case_when(
      str_trim(str_to_lower(Sex)) == "male" ~ 1,
      str_trim(str_to_lower(Sex)) == "female" ~ 0,
    ),
    
    # Encode income_category into an ordinal numeric scale (1-3).
    # This reflects the 'high/medium/low' structure described in the brief
    # and allows the model to treat income as a continuous predictor if desired.
    income_num = case_when(
      str_trim(str_to_lower(income_category)) == "low" ~ 1,
      str_trim(str_to_lower(income_category)) == "medium" ~ 2,
      str_trim(str_to_lower(income_category)) == "high" ~ 3,
    )
  )
```


According to the task, our objective is to examine how all variables influence food expenditure. As part of the EDA, we generate bivariate scatterplots to explore the relationship between each predictor—both numerical and categorical—and food expenditure. Scatterplots are an essential component of this process because they allow us to visually assess the form, direction, and strength of these relationships in a way that summary statistics alone cannot capture. They also help reveal patterns such as linearity, clustering, or potential non‑linear behaviour, while making it easier to detect unusual observations that may affect model performance. These visual insights guide our understanding of the data structure before moving into formal regression modelling, with the important reminder that correlation does not imply causation.

```{r}
# Scatterplots of Food vs Alcohol, Debt, and Housing Expenditure

# Food vs Alcohol
ggplot(householdmain_clean, aes(x = Alcohol, y = Food)) +
  geom_point(alpha = 0.6, colour = "#E67E22") +
  geom_smooth(method = "lm", colour = "#2C3E50") +
  labs(
    title = "Scatterplot of Food vs. Alcohol Expenditure",
    x = "Weekly Alcohol Expenditure (£)",
    y = "Weekly Food Expenditure (£)"
  ) +
  theme_minimal()

# Food vs Debt
ggplot(householdmain_clean, aes(x = Debt, y = Food)) +
  geom_point(alpha = 0.6, colour = "#27AE60") +
  geom_smooth(method = "lm", colour = "#2C3E50") +
  labs(
    title = "Scatterplot of Food vs. Debt Expenditure",
    x = "Weekly Debt Payments (£)",
    y = "Weekly Food Expenditure (£)"
  ) +
  theme_minimal()

# Food vs Housing
ggplot(householdmain_clean, aes(x = Housing, y = Food)) +
  geom_point(alpha = 0.6, colour = "#8E44AD") +
  geom_smooth(method = "lm", colour = "#2C3E50") +
  labs(
    title = "Scatterplot of Food vs. Housing Expenditure",
    x = "Weekly Housing Expenditure (£)",
    y = "Weekly Food Expenditure (£)"
  ) +
  theme_minimal()
```


The scatterplots provide a visual assessment of the relationship between food expenditure and each predictor variable. In the plot of Food vs. Alcohol, we observe a slight positive trend, suggesting that households spending more on alcohol tend to spend marginally more on food. The fitted line reinforces this directionality, though the spread of points indicates moderate variability.

In contrast, the Food vs. Debt scatterplot shows a dense central cluster with no clear upward or downward trend, implying a weak or negligible correlation between debt payments and food expenditure. The wide dispersion of points and flat regression line suggest that debt levels may not be a strong predictor of food spending.

The Food vs. Housing plot displays a clearer positive relationship, with food expenditure increasing alongside housing costs. The fitted line and confidence band indicate a more consistent trend, suggesting that households with higher housing expenses may also allocate more towards food, potentially reflecting overall affluence or cost-of-living effects.

These visual patterns help guide variable selection and model specification, while reinforcing the principle that correlation does not imply causation. Further statistical testing is required to validate these associations.

Next we construct the box plot for the nominal categorical variables(sex and income_category) to understand the effect of the variables on food expenditure

```{r}
## Boxplot: Food by Income Category
ggplot(householdmain_clean, aes(x = income_category, y = Food, fill = income_category)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Food Expenditure by Income Category",
    x = "Income Category",
    y = "Weekly Food Expenditure (£)"
  ) +
  theme_minimal()

## Boxplot: Food by Sex
ggplot(householdmain_clean, aes(x = Sex, y = Food, fill = Sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Food Expenditure by Sex",
    x = "Sex",
    y = "Weekly Food Expenditure (£)"
  ) +
  theme_minimal()
```


The boxplots for Sex and Income Category provide a visual comparison of food expenditure across categorical groups. For Sex, both male and female groups exhibit similar medians and interquartile ranges, suggesting no substantial difference in typical food spending between genders. However, the presence of multiple outliers in both groups indicates variability in individual spending habits.

In the case of Income Category, the distributions across Low, Medium, and High income groups show comparable medians and spreads, with no strong upward trend in food expenditure as income increases. This suggests that food spending may not be directly proportional to income level. Outliers are present in all categories, indicating that some households—regardless of income—spend significantly more or less on food than the typical range.

These patterns highlight that categorical variables like Sex and Income Category may have limited explanatory power for food expenditure on their own, and should be evaluated alongside other predictors in the regression analysis.

To further explore how household structure relates to food expenditure, jitter plots are used for the ordinal variables Kids, HouseholdSize, and Vehicles. Because these variables take on a limited set of discrete values, jittering helps reveal the spread of food spending within each category and makes it easier to see whether expenditure tends to rise, fall, or remain stable as these counts increase. This provides a clearer sense of how variation in household composition may align with differences in food spending patterns.

```{r}
# Boxplot + jitter for kids (treating number of kids as categorical)
householdmain_clean %>%
  mutate(kids_f = factor(Kids,levels = sort(unique(Kids)))) %>%
  ggplot(aes(x = kids_f, y = Food)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, fill = "grey90") +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(x = "Number of kids", y = "Weekly food expenditure (£)", title = "Food expenditure by number of kids") +
  theme_minimal()

# Boxplot + jitter for HouseholdSize (treating Size of the Household as categorical)
householdmain_clean %>%
  mutate(household_f = factor(HouseholdSize, levels = sort(unique(HouseholdSize)))) %>%
  ggplot(aes(x = household_f, y = Food))+
  geom_boxplot(outlier.shape = NA, width = 0.6, fill = "grey90") +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(x = "Household size", y = "Weekly food expenditure (£)", title = "Food expenditure by household size") +
  theme_minimal()

# Boxplot + jitter for vehicles (treating number of vehicles as categorical)
householdmain_clean %>%
  mutate(vehicles_f = factor(Vehicles, levels = sort(unique(Vehicles)))) %>%
  ggplot(aes(x = vehicles_f, y = Food)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, fill = "grey90") +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(x = "Number of vehicles", y = "Weekly food expenditure (£)", title = "Food expenditure by number of vehicles") +
  theme_minimal()
```


The jitter plots reveal how food expenditure varies across discrete levels of Kids, HouseholdSize, and Vehicles. For Kids, there is a visible upward shift in both median and mean expenditure as the number of children increases, suggesting a positive association between family size and food costs. Similarly, HouseholdSize shows a gradual rise in expenditure up to size 6–8, beyond which the sample size becomes sparse and variability increases. The Vehicles plot shows a less pronounced pattern, with food expenditure remaining relatively stable across most categories, though households with 5 or more vehicles exhibit slightly higher spending.

These visual patterns suggest that Kids and HouseholdSize may be stronger predictors of food expenditure than Vehicles, and they provide useful guidance for variable selection in subsequent regression modelling.

The scatterplots and jitter plots collectively offer a visual overview of how both continuous and ordinal variables relate to food expenditure, revealing patterns of association, variability, and potential predictors. While some variables show clear positive trends (e.g., housing cost, number of kids), others exhibit weak or inconsistent relationships, reinforcing the need for formal modelling to validate these observations.

### Multiple Regression Analysis

Multiple regression analysis provides several outputs that help us assess how each predictor contributes to weekly food expenditure. The regression coefficients quantify the expected change in food spending for a one‑unit increase in each predictor, while holding all other variables constant. These estimates allow us to evaluate how household characteristics—such as housing expenditure, number of children, income category, or alcohol spending—are associated with variation in food expenditure.

The model also reports NHST‑based statistics, including p-values and confidence intervals, which indicate whether these associations are statistically reliable. This enables us to identify which predictors have meaningful explanatory value and which do not.

In addition, regression allows us to test interaction effects, which are central to the question of whether the impact of alcohol expenditure differs across households with varying levels of debt. By incorporating an alcohol × debt interaction term, the model can determine whether the relationship between alcohol spending and food expenditure changes depending on debt payments, revealing potential moderation effects.

Taken together, the regression coefficients, significance tests, confidence intervals, and interaction estimates provide a robust framework for understanding how household characteristics shape food expenditure and for communicating these findings clearly and accurately.

```{r}
# --- Step 1: Data Preparation ---
householdmain_clean <- householdmain_clean |>
  mutate(
    Kids = factor(Kids),
    HouseholdSize = factor(HouseholdSize),
    Vehicles = factor(Vehicles)
  )

# --- Step 2: Run Main Model ---
model_main <- lm(
  Food ~ HouseholdSize + Kids + Vehicles + Housing +
    Alcohol + Debt + sex_num + income_num,
  data = householdmain_clean
)

# --- Step 3: Run Main Model ---
model_interaction <- lm(
  Food ~ HouseholdSize + Kids + Vehicles + Housing +
    Alcohol * Debt + sex_num + income_num,
  data = householdmain_clean
)
```


```{r}
# Automatically detect all factor levels and create clean names
kids_levels <- levels(householdmain_clean$Kids)
household_levels <- levels(householdmain_clean$HouseholdSize)
vehicle_levels <- levels(householdmain_clean$Vehicles)

coef_names <- c(
  "(Intercept)" = "Constant (Intercept)",
  "Housing"     = "Housing Expenditure (£)",
  "Alcohol"     = "Alcohol Expenditure (£)",
  "Debt"        = "Debt Payments (£)",
  "sex_num"     = "Sex (Male)",
  "income_num"  = "Income Level (1–3)",
  "Alcohol:Debt" = "Alcohol × Debt Interaction",

  # Kids factor levels
  setNames(
    paste0("Kids: ", kids_levels),
    paste0("Kids", kids_levels)
  ),

  # Household size factor levels
  setNames(
    paste0("Household Size: ", household_levels),
    paste0("HouseholdSize", household_levels)
  ),

  # Vehicles factor levels
  setNames(
    paste0("Vehicles: ", vehicle_levels),
    paste0("Vehicles", vehicle_levels)
  )
)

#Publishes the table summarizing the NHST and estimation values for main and interaction model
models <- list(
  "Model 1: Main Effects" = model_main,
  "Model 2: Interaction"  = model_interaction
)

modelsummary(
  models,
  coef_map = coef_names,
  statistic = c(
    "Beta = {estimate}",
    "Std. Error = {std.error}",
    "t = {statistic}",
    "p = {p.value}",
    "95% CI = [{conf.low}, {conf.high}]"
  ),
  shape = term + statistic ~ model,
  stars = TRUE,
  gof_omit = "AIC|BIC|Log.Lik|F",
  title = "Regression Analysis Predicting Weekly Food Expenditure",
  notes = "Rows show Beta, Standard Error, t-statistic, p-value, and 95% CI for each predictor.",
  output = "kableExtra"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) |>
  column_spec(1, bold = TRUE, width = "5cm")
```

### Technical Interpretation of Regression Values

Both models examine how various economic and demographic factors predict weekly food expenditure.

#### Significant Economic Predictors:

- Housing Expenditure (£): In both models, housing spend is a highly significant predictor ($\beta = 0.092, p < 0.001$). For every £1 increase in housing costs, weekly food expenditure is estimated to increase by approximately 9.2 pence. The narrow confidence interval $[0.085, 0.099]$ indicates high precision in this estimate.

- Alcohol Expenditure (£): This has a significant positive relationship with food spend. In Model 1, every £1 spent on alcohol predicts a 39.4p increase in food spend ($\beta = 0.394, p < 0.001$).Debt Payments (£): In the main effects model, debt has a significant negative impact ($\beta = -0.051, p < 0.001$). Every £1 of debt payment reduces food expenditure by approximately 5 pence, acting as a financial constraint.

#### Demographic Factors:

- Household Size: Large households specifically (Size 8) show a significant increase in food spend ($\beta = 14.48, p = 0.005$), adding roughly £14.48 to the weekly bill compared to a single-person household.

- Number of Children (Kids): Interestingly, very large numbers of children (6 and 8) show significant negative coefficients (e.g., -£36.70 for 8 kids, $p = 0.030$). This may indicate significant "economies of scale" or budget substitution in extremely large families.

- Vehicles: Owning 2 vehicles is significantly associated with an increase of £1.16 in weekly food spending ($p = 0.031$).

- Interaction Analysis (Model 2):The Alcohol × Debt Interaction was not statistically significant ($\beta = -0.002, p = 0.597$). This suggests that the relationship between alcohol spending and food spending does not change significantly based on the level of debt a household carries.

In simpler terms, the data tells a clear story about what drives a household's food budget:

 - Households that spend more on housing and alcohol also spend more on food. Specifically, for every extra £100 spent on housing, food costs go up by about £9.20. This suggests that food spending scales with overall household wealth and lifestyle.

- Debt acts as a direct competitor to the dinner table. As debt payments increase, families are forced to cut back on their food budget. Every £20 of debt repayment removes about £1 from the weekly grocery shop.

- While larger households (8 people) naturally spend more on food (about £14.50 more per week), having a very high number of children actually correlates with lower spending. This likely reflects the financial pressure of very large families, who may be forced to buy in bulk or choose cheaper, more cost-effective food options to survive.

- The interaction test showed that debt doesn't really change how alcohol spend relates to food spend. Whether a family has high or low debt, their habit of spending more on food when they spend more on alcohol remains consistent.

- Interestingly, a person's Sex or their specific Income Level (1–3) did not have a statistically significant impact on food spending once other factors like housing and household size were accounted for.

```{r}
# --- Step 4: Visualizations ---

# 1. Coefficient Plot 
model_df <- tidy(model_main, conf.int = TRUE) |>
  filter(term != "(Intercept)")



ggplot(model_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(color = "steelblue", size = 3) +
  # Using geom_errorbar
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
                width = 0.2, 
                orientation = "y", 
                color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Estimated Predictors of Food Expenditure",
       subtitle = "Coefficient estimates with 95% Confidence Intervals",
       x = "Beta Coefficient (Standardized)", 
       y = "Predictor") +
  theme_minimal(base_size = 14)


# 1. Prepare the data from both models
# We combine them so you can see if significance changed between Model 1 and Model 2
heatmap_data <- bind_rows(
  tidy(model_main) |> mutate(model = "Model 1: Main Effects"),
  tidy(model_interaction) |> mutate(model = "Model 2: Interaction")
) |>
  filter(term != "(Intercept)") |>
  mutate(
    # Clean up variable names for the plot
    term = case_when(
      term == "sex_num" ~ "Sex (Male)",
      term == "income_num" ~ "Income Level",
      term == "Alcohol:Debt" ~ "Alcohol × Debt",
      TRUE ~ term
      
    ),
    # Create significance categories for easier reading
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
    )
  )

# 2. Create the Heatmap
ggplot(heatmap_data, aes(x = model, y = reorder(term, p.value), fill = p.value)) +
  geom_tile(color = "white", lwd = 0.8) +
  # Add the significance stars as text on top of the tiles
  geom_text(aes(label = significance), color = "black", size = 5) +
  # Use a professional color gradient (Low p-values = dark/significant)
  scale_fill_gradient(low = "#43a2ca", high = "#e0f3db", name = "p-value") +
  labs(
    title = "Heatmap of Predictor Significance",
    subtitle = "Comparing Main Effects vs. Interaction Model",
    x = "Model Version",
    y = "Predictor Variable",
    caption = "Stars indicate significance: *<0.05, **<0.01, ***<0.001"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(),
    axis.text.x = element_text(angle = 0)
  )


# 3. Interaction Plot
interaction_data <- ggpredict(model_interaction, terms = c("Alcohol", "Debt"))



plot(interaction_data) +
  labs(title = "Interaction Effect of Alcohol and Debt",
       subtitle = "Predicted food spending at varying levels of alcohol and debt",
       x = "Weekly Alcohol Expenditure (£)",
       y = "Predicted Food Expenditure (£)",
       colour = "Debt Level") +
  theme_minimal(base_size = 14)
```


### Inference of Plot Visualizations of the NHST and Estimation Approaches 

#### Figure 1 — Standardised beta coefficients (Estimated Predictors of Food Expenditure)
The plot displays standardised effect sizes and 95% confidence intervals for each predictor, allowing direct comparison of relative importance. Housing and Alcohol have the largest positive standardised coefficients with CIs that do not cross zero, indicating they are the most robust positive correlates of food expenditure; several factor levels for household size and kids show large point estimates but wide CIs, signalling imprecision. Predictors whose intervals cross the vertical zero line provide no reliable evidence of an effect in these data.

Larger household sizes (specifically households with 8, 9, or 11 people) are the strongest predictors of higher food spending. This makes sense—more mouths to feed equals a higher bill.Interestingly, having more children (labeled as Kids 5, 6, and 8) seems to correlate with lower food expenditure in this specific data. This might suggest that as families get very large, they may become more budget-conscious or buy in bulk to save money.Factors like income, sex, and having just 1 or 2 kids have very small dots near the zero line, meaning they don't change food spending as much as you might expect.

#### Figure 2 — Heatmap of predictor significance (Model 1 vs Model 2)
The heatmap encodes p‑values across the two model specifications, with darker cells and stars marking stronger statistical evidence. It highlights that Alcohol is consistently significant in both models, Debt is significant only in the main‑effects specification, and HouseholdSize = 8 emerges as significant in the interaction model; many other predictors show weak or inconsistent significance across specifications. This visual summary emphasises which variables warrant substantive attention and which results are sensitive to model choice.

Housing, Alcohol, and Debt are highly significant (marked with ***). We can be very confident these factors play a role.When the researchers moved from a simple model (Model 1) to a more complex one (Model 2), Debt lost its significance. This suggests that debt itself might not be the direct cause of spending changes, but rather how it interacts with other life factors.

#### Figure 3 — Interaction plot: Alcohol × Debt (Predicted Food Expenditure)
The interaction plot shows predicted weekly food expenditure across a range of alcohol spending for three representative debt levels, with shaded confidence bands. All lines slope upward (higher alcohol → higher predicted food spend), but the slopes and intercepts are similar and confidence bands overlap substantially, visually corroborating the regression result that the alcohol–food relationship does not differ meaningfully by debt level.

Across the board, as people spend more on alcohol (moving right on the horizontal axis), their predicted food spending also goes up.People with lower debt (the red line) actually spend the most on food as they increase their alcohol spending.People with higher debt (the green line) spend the least on food overall.

High debt seems to "suppress" food spending. Even if a high-debt household increases their alcohol budget, they still spend less on food than a low-debt household doing the same thing.

### Interpretation of ANOVA Table

The nested‑model ANOVA comparing the main‑effects specification (Model 1) to the interaction specification (Model 2) indicates that adding the Alcohol × Debt term does not materially improve model fit. Numerically, Model 1 has residual df = 4054 and RSS = 574,880, while Model 2 (one additional parameter) has residual df = 4053 and RSS = 574,841, yielding a sum‑of‑squares reduction of 39.71 for 1 degree of freedom. The corresponding F‑test is F(1, 4053) = 0.28 with p = 0.597, which is far above conventional significance thresholds. In practical terms, the interaction explains a negligible amount of additional variance and the evidence is insufficient to justify the more complex model; therefore the simpler main‑effects model is preferred on both parsimony and statistical grounds. For completeness, one might still report the interaction estimates and confidence intervals, and consider alternative specifications (e.g., centred predictors or non‑linear terms) if there is a substantive reason to suspect moderation that the current linear interaction fails to capture.

```{r}
#Creating a summary of Anova analysis table
anova(model_main, model_interaction)
```


The ANOVA‑derived diagnostic plots reinforce the numeric test that the Alcohol × Debt interaction does not meaningfully improve model fit. The estimated marginal‑means plot shows the modelled average food‑expenditure margins for the Alcohol and Debt strata with their 95% confidence intervals; the intervals overlap substantially, indicating no clear separation between groups. The interaction panel (predicted food expenditure at a representative debt level) likewise displays a central point estimate with wide vertical error bars, and the shaded confidence bands in the interaction line plot overlap across debt strata—visually mirroring the small sum‑of‑squares reduction and non‑significant F‑test (F(1,4053) ≈ 0.28, p ≈ 0.60). Taken together, these ANOVA plots provide graphical confirmation that adding the linear Alcohol × Debt term yields negligible explanatory gain and that the simpler main‑effects specification is preferable on parsimony and statistical grounds.

The results of this comparison indicate that the extra complexity is not necessary. While Model 2 explains a tiny bit more variation in the data (represented by the "Sum of Sq" value of 39.712), the p-value of 0.5967 is the deciding factor. Because this value is much higher than the standard threshold of 0.05, researchers conclude that the improvement is purely due to chance rather than a real-world effect. The simpler Model 1 is just as effective at predicting food spending as the more complicated Model 2, and the interaction between alcohol and debt does not offer meaningful new insight into the data.


```{r}
library(emmeans)
library(ggplot2)

# Get estimated marginal means for alcohol at different debt levels
emm <- emmeans(model_interaction, ~ Alcohol | Debt)

# Plot the marginal means
plot(emm)  # quick base R plot

# Or use ggplot for publication-ready version
emm_df <- as.data.frame(emm)

ggplot(emm_df, aes(x = Alcohol, y = emmean, color = factor(Debt))) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  labs(
    title = "Interaction: Alcohol × Debt",
    x = "Alcohol Spending",
    y = "Predicted Food Expenditure",
    color = "Debt Level"
  ) +
  theme_minimal()

```

The two graphs illustrate why the complex model failed to provide new insights.

- The Interaction Plot: Shows a single data point for food expenditure (${\approx}54$) at a specific level of alcohol and debt. The vertical red line is very wide, ranging from roughly $48$ to $59$. This indicates there is too much uncertainty or "noise" to claim a specific interaction effect exists.

- EM-Mean Plot: This is just another way of looking at that same point. It shows that at an alcohol spending level of roughly $17.02$ and a debt level of $59.98$, the average predicted food expenditure is approximately $53.8$. The long light-blue bar represents the same wide range of uncertainty.

We perform Residual analysis to assess whether the linear regression model meets core assumptions—linearity, normality of errors, homoscedasticity, and absence of influential observations—so that coefficient estimates, standard errors and hypothesis tests are valid.

The script augments the analysis dataset with model diagnostics (fitted, residuals, standardized residuals, and Cook’s distance) and produces four diagnostic plots: Residuals vs Fitted, Normal Q–Q, Scale‑Location, and Cook’s Distance. An optional 2×2 diagnostic panel (plot(model_main)) and performance::check_model() are available for a compact summary.

Diagnostic plots and interpretation

Residuals vs Fitted evaluates linearity and omitted structure. A horizontal cloud with a near‑flat LOESS line supports the linear specification; systematic curvature or patterns would indicate model misspecification or omitted nonlinear terms.

Normal Q–Q assesses error normality. Points close to the reference line indicate approximate normality; systematic tail deviations suggest skewness or heavy tails that may affect inference in small samples.

Scale‑Location inspects homoscedasticity by plotting standardized residuals against fitted values. A flat trend supports constant variance; an upward or downward trend indicates heteroscedasticity and motivates robust standard errors or variance‑stabilising transforms.

Cook’s Distance identifies influential observations. Observations exceeding the conventional threshold 
4
/
𝑛
 should be inspected; report their case IDs and assess sensitivity of coefficients to their exclusion.

```{r}
# Full diagnostic panel (2x2) for residual analysis
par(mfrow = c(2, 2))
plot(model_main)
par(mfrow = c(1, 1))
```
The diagnostic panel indicates the model is broadly well behaved but not perfect. The Residuals vs Fitted plot shows a roughly symmetric cloud of points around zero with no pronounced systematic curvature, suggesting the linear specification captures the main trends and there is no obvious omitted nonlinear pattern. The Normal Q–Q plot shows residuals close to the reference line for most of the distribution, with only modest deviation in the tails, implying approximate normality adequate for large‑sample inference but caution for tail‑sensitive tests. The Scale‑Location plot reveals a mild upward trend in standardized residuals across fitted values, indicating some heteroscedasticity; this is not extreme but justifies reporting heteroscedasticity‑robust standard errors. The Residuals vs Leverage (Cook’s) panel identifies a small number of observations with elevated influence; these cases should be inspected and a sensitivity check reported (re‑estimate coefficients with and without flagged observations).

Model assumptions are sufficiently met for interpretation of coefficients, with two caveats — mild heteroscedasticity (use robust SEs) and a few potentially influential observations (report sensitivity analyses).



### Conclusions and inference

##### Summary of key findings

The analysis identifies housing expenditure and alcohol expenditure as the most consistent positive correlates of weekly food spending. Housing shows a highly precise positive effect (unstandardised β = 0.092, t ≈ 26.33, p < .001). Alcohol spending is also positively associated with food spending (Model 1 β = 0.394, p < .001; Model 2 β = 0.493, p = .012). The Alcohol × Debt interaction is small and statistically non‑significant (β = −0.002, t = −0.53, p = .597), and the nested ANOVA confirms the interaction does not improve fit (F ≈ 0.28, p ≈ 0.60). Several categorical levels for household size, number of kids and vehicles produce large point estimates but have wide confidence intervals and should not be over‑interpreted.

##### Model performance and reliability

The model explains a modest share of variance in weekly food expenditure (R² = 0.158, adjusted R² ≈ 0.153) and has a typical prediction error of RMSE = £11.87. This indicates the identified associations are statistically reliable for population‑level inference, but the model has limited precision for household‑level prediction.

##### Diagnostic assessment

Residual diagnostics show the model is broadly well behaved: the Residuals vs Fitted plot shows no pronounced nonlinearity, the Q–Q plot indicates approximate normality with minor tail deviations, and the Scale‑Location plot suggests mild heteroscedasticity. The influence plot flags a small number of potentially influential observations. Taken together, these diagnostics justify reporting heteroscedasticity‑robust standard errors and performing sensitivity checks that exclude flagged cases.

##### Practical implications

- Targeting and segmentation: Households with higher housing and alcohol expenditures are likely to spend more on food and represent priority segments for cross‑sell or premium grocery offers.

- Debt considerations: There is no evidence that household debt materially moderates the alcohol–food relationship in this linear specification; strategies predicated on such moderation are not supported by these results.

- Caution on sparse categories: Decisions based on extreme household sizes or rare category levels should be deferred until validated with larger samples.

Overall, the evidence supports housing and alcohol as the principal correlates of weekly food expenditure in these data, while the hypothesised moderation by debt is not supported; model diagnostics warrant modest corrective steps (robust SEs and sensitivity checks) but do not undermine the principal substantive conclusions.

# Question 2

The objective of this analysis is to evaluate the primary drivers of podcast popularity across various streaming platforms by examining a dataset where each observation represents a unique podcast's average performance over several months. The investigation focuses on three core areas: the relationship between daily.listens and the combined influence of avg.review scores and total.reviews volume, a contrast of the distribution of listens across different categories to identify genre-specific performance standards, and the use of interaction modeling to determine if the effect of subscribers on daily.listens is moderated by the podcast’s genre. By applying multiple linear regression and ANOVA—validated through rigorous residual diagnostics—this report identifies which metrics most reliably predict success in the current podcasting market.

### Data Dictionary

The variables from the data provided are described in the table below

| Variable | Description |
|------------------------------------|------------------------------------|
| Platform | The streaming service/platform where the podcast is hosted. |
| Genre | The specific category or theme of the podcast content. |
| avg.review | The average review score received by the podcast. |
| daily.listens | The total volume of reviews submitted for the podcast. |
| total.reviews | The average number of listens across all days in the period. |
| subscribers | The total number of users subscribed to the podcast. |

To begin the analysis, the dataset "podcast_listens.csv" is imported and stored as a tibble named "podcast_listens"

```{r}
podcast_listens <- read_csv("podcast_listens.csv")
```
### Data Dictionary

The variables from the data provided are described in the table below

| Variable | Description |
|------------------------------------|------------------------------------|
| Platform | The streaming service/platform where the podcast is hosted. |
| Genre | The specific category or theme of the podcast content. |
| avg.review | The average review score received by the podcast. |
| daily.listens | The total volume of reviews submitted for the podcast. |
| total.reviews | The average number of listens across all days in the period. |
| subscribers | The total number of users subscribed to the podcast. |


It comprises 5400 observations across 6 variables. The column types are as follows: two character variables (Platform, genre) and four numeric variables (avg.review, daily.listens, total.reviews, subscribers). This structure supports subsequent analysis of listening patterns, platform performance, and genre-based trends.

We summarised the dataset using R’s summary() function to report the minimum, 1st quartile, median, mean, 3rd quartile and maximum for each numeric variable; standard deviations and counts of missing values were computed separately (e.g., sd() and sapply(..., function(x) sum(is.na(x)))) and IQR‑based checks were used to flag potential outliers. The resulting table and visual checks (boxplots, histograms) were used to identify implausible values, variables with high dispersion, and any columns with non‑negligible missingness for further cleaning or sensitivity analysis.

```{r}
summary(podcast_listens)
```
D
+++++++aily.listens and total.reviews show reasonable, tight distributions; avg.review has an extreme max (100) and subscribers contains an implausible negative value (−2), both requiring inspection/correction.

We identified two categorical variables—Platform and Genre—and examined their level distributions to assess sample balance, detect sparse categories, and inform modelling choices.

```{r}
cat("\nDistribution of Platform:\n") 
table(podcast_listens$Platform) 

cat("\nDistribution of Genre:\n") 
table(podcast_listens$genre)
```

A new tibble, "Podcast_numeric", was created containing only the numeric variables to streamline descriptive statistics, outlier detection, and subsequent quantitative analyses.

```{r}
# Select only expenditure variables
(Podcast_numeric <- podcast_listens %>%
  select(avg.review,daily.listens,total.reviews,subscribers))
```

Histograms were plotted for daily_listens, subscribers, total_reviews, and avg_reviews to visually assess distribution shape, skewness, multimodality, and potential outliers

```{r}
Podcast_numeric %>%
  # Keep only the numeric columns to ensure the plotting function doesn't error
  select(where(is.numeric)) %>%
  
  # Exclude 'avg.review' as it likely has a different scale or isn't needed for this visual
  select(-avg.review) %>%  
  
  # Reshape data from "wide" to "long" format so all variables can be plotted at once
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  
  # Initialize the plot using 'value' for the x-axis
  ggplot(aes(x = value)) +
  
  # Create histograms with specific styling (blue fill, white borders, and 60 bars)
  geom_histogram(
    fill = "#2C7FB8", color = "white", bins = 60, alpha = 0.85
  ) +
  
  # Create a separate small plot for each variable; 'scales = free' lets each x-axis adjust
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  
  # Add descriptive titles and label the axes
  labs(
    title = "Distribution of Numeric Podcast Metrics",
    subtitle = "Histograms for listens, reviews, subscribers",
    x = "Value",
    y = "Count"
  ) +
  
  # Apply a clean, modern theme and customize font sizes/colors for better readability
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#08306B"),
    plot.subtitle = element_text(size = 12, color = "#08519C"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85")
  )
```

```{r}
Podcast_numeric %>%
  # Isolate only the 'avg.review' column for this specific plot
  select(avg.review) %>%
  
  # Initialize the plot with 'avg.review' on the x-axis
  ggplot(aes(x = avg.review)) +
  
  # Create a histogram using a high number of bins (100) to show fine details in the scores
  geom_histogram(
    fill = "#2C7FB8",  # Set bar color to a professional blue
    color = "white",   # Add white borders for better separation between bars
    bins = 100,        # Use a high bin count to capture small variations in review averages
    alpha = 0.85       # Set slight transparency for a cleaner look
  ) +
  
  # Add descriptive titles and label the axes
  labs(
    title = "Distribution of Average Review Scores",
    x = "Average Review Score",
    y = "Count"
  ) + 
  
  # Force the x-axis range from -5 to 6 to provide context (or see if there are outliers)
  xlim(-5, 6) + 
  
  # Apply a clean, modern theme with a base font size of 13
  theme_minimal(base_size = 13) +
  
  # Custom styling for the title and grid lines to improve readability
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#08306B"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),           # Remove minor grid lines to reduce clutter
    panel.grid.major = element_line(color = "grey85") # Soften major grid lines
  )
```

The histograms show daily_listens is unimodal and concentrated around the mid‑4000s, while subscribers and total_reviews are right‑skewed with long upper tails, indicating a few high‑value observations influence means. avg_reviews is tightly clustered near typical rating values but contains implausible extremes that require validation.

Boxplots were generated for all numerical variables to identify the precise locations of potential outliers. Observations falling beyond the 1.5×IQR whiskers were flagged for further inspection, allowing a clear distinction between typical variation and extreme values that may require additional validation or treatment in subsequent analysis.

```{r}
Podcast_numeric %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  
  # Boxplot styling
  geom_boxplot(
    fill = "#74A9CF",
    color = "#045A8D",
    width = 0.6,
    outlier.color = "#023858",
    outlier.size = 1.8,
    alpha = 0.9
  ) +
  
  # Facets
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  
  # Labels
  labs(
    title = "Boxplots of Numeric Variables",
    subtitle = "Visual inspection of distribution and outliers across all numeric fields",
    x = "",
    y = "Value"
  ) +
  
  # Theme enhancements
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16,
      color = "#08306B"
    ),
    plot.subtitle = element_text(
      size = 12,
      color = "#08519C"
    ),
    strip.text = element_text(
      face = "bold",
      size = 12,
      color = "white"
    ),
    strip.background = element_rect(
      fill = "#045A8D",
      color = NA
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    panel.border = element_rect(
      fill = NA,
      color = "grey70",
      size = 0.5
    )
  )
```


The boxplots reveal distinct distribution patterns and outlier behaviour across the four numeric podcast metrics:

Average Review Score: The distribution is tightly concentrated between 3 and 5, indicating generally positive feedback. However, a significant outlier above 100 suggests a likely data entry error that should be flagged for correction or exclusion.

Daily Listens: Most values fall within a narrow range around 4400–4600, with a few outliers exceeding 5000. This reflects consistent audience engagement with occasional spikes, possibly due to promotional boosts or viral episodes.

Subscribers: The central tendency lies around 600–700, but outliers above 1200 indicate a subset of podcasts with unusually large followings. These may represent high-profile or long-running shows and warrant further segmentation analysis.

Total Reviews: The distribution is centered near 1000, with multiple outliers above 1500. These high-review counts suggest elevated listener interaction, which may correlate with popularity or longevity.

To improve data quality and ensure meaningful analysis, we removed the extreme value of 100 from the avg.review variable, as it falls well outside the plausible rating range and likely reflects a data entry error. Additionally, we retained only observations with subscribers > 0, excluding zero values that may represent inactive or placeholder entries rather than genuine audience metrics. These cleaning steps help reduce distortion in summary statistics and improve the reliability of downstream modelling.

After the initial outlier inspection, each variable in the podcast_listens dataset was assessed for missing values. Any observations containing NA entries will be removed to ensure that subsequent descriptive analysis and modelling were based on complete and reliable records. This step prevents biased estimates and avoids complications arising from incomplete cases during statistical procedures.

```{r}
# --- Load Library ---
library(tidyverse)
library(knitr)

# --- Check for Missing Values ---
# We calculate the sum of NA values for each column
(na_summary <- podcast_listens %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "NA_Count"))
```

A summary check for missing values across all key variables in the podcast_listens dataset confirmed that no NA entries were present in any column, including Platform, genre, avg.review, daily.listens, total.reviews, and subscribers. This indicates a fully complete dataset with no gaps in observation, allowing for uninterrupted descriptive analysis and modelling without the need for imputation or case-wise deletion. The absence of missing data strengthens the reliability of statistical inference and ensures consistency across visualizations and model inputs.

```{r}
(podcast_clean <- podcast_listens %>%
  filter(avg.review <= 5, subscribers >= 0) %>%
  mutate(
    genre = relevel(as.factor(genre), ref = "Comedy"),
    Platform = as.factor(Platform)
  ))
```

To ensure data integrity 2 and model validity, the podcast_listens dataset was cleaned as follows:

- Logical Filtering: avg.review was restricted to $\le 5$ to remove out-of-range errors, and subscribers was filtered to $\ge 0$ to eliminate impossible negative values.

- Factor Conversion: genre and Platform were converted to factors. This follows standard R conventions for categorical data, allowing the lm() function to automatically generate necessary dummy variables

- Baseline Selection: The relevel() function was used to set "Comedy" as the reference level for genre. This ensures all subsequent regression coefficients represent the change in listens relative to Comedy podcasts, providing a clear baseline for comparison.

### Relationship Between Reviews and Listenership

To investigate if podcast listens vary based on average review scores and total review volume, a multiple linear regression analysis was performed.

Visual Exploration: A color-coded scatter plot was generated to map three variables simultaneously: avg.review (x-axis), daily.listens (y-axis), and total.reviews (color gradient). This provides an immediate visual check of whether high-volume or high-scoring podcasts dominate listenership.

Statistical Modeling: A linear model (lm) was fitted using avg.review and total.reviews as predictors for daily.listens.

Assumption Validation: Diagnostic residual plots (Residuals vs Fitted, Normal Q-Q, Scale-Location, and Residuals vs Leverage) were generated to verify the linear regression assumptions of linearity, normality, and homoscedasticity.

Effect Visualization: A Coefficient Forest Plot was created to display the beta estimates and their 95% confidence intervals. This facilitates an "estimation approach" by showing the precision and direction of each predictor's effect relative to zero.

```{r}
# Visualization: 3-Variable Relationship
ggplot(podcast_clean, aes(x = avg.review, y = daily.listens, color = total.reviews)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") + # General trend line
  scale_color_viridis_c() + # Professional, colorblind-friendly scale
  labs(
    title = "Podcast Listens by Review Score and Review Volume",
    x = "Average Review Score",
    y = "Daily Listens",
    color = "Total Reviews"
  ) +
  theme_minimal()

# Multiple Linear Regression
model_1 <- lm(daily.listens ~ avg.review + total.reviews, data = podcast_clean)

# Check assumptions (Linearity, Normality of residuals, Homoscedasticity)
par(mfrow = c(2, 2))
plot(model_1)
title("Diagnostic Plots for Model 1", outer = TRUE)


# Summary of results
summary(model_1)

# 1. Load necessary libraries
library(tidyverse) # For data manipulation and plotting
library(broom)     # For tidying regression results

# 2. Tidy the model and calculate 95% Confidence Intervals
# We filter out the '(Intercept)' because its scale is usually much larger 
# than the slopes, which can squash the visual representation of other variables.
model_1_tidy <- tidy(model_1, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

# 3. Create the Forest Plot
ggplot(model_1_tidy, aes(x = estimate, y = term)) +
  # Adds the point estimate
  geom_point(size = 3, color = "#2c3e50") +
  # Adds the horizontal error bars (95% Confidence Intervals)
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#34495e") +
  # Adds a vertical reference line at zero
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  # Professional labels and theme
  labs(
    title = "Coefficient Forest Plot: Predictors of Daily Listens",
    subtitle = "Points represent estimates; error bars represent 95% Confidence Intervals",
    x = "Regression Coefficient (Estimate)",
    y = "Variable",
    caption = "Note: Red dashed line at 0 indicates no effect."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "italic")
  )
```


A multiple linear regression was conducted to examine the impact of avg.review and total.reviews on podcast daily.listens. The overall model was statistically significant, $F(2, 5373) = 80.24, p < 2.2 \times 10^{-16}$, although the adjusted $R^{2}$ of $0.02864$ indicates that reviews explain only approximately $2.9\%$ of the variance in listens.

- Total Reviews: There is a highly significant positive relationship between the volume of reviews and daily listens ($\beta = 0.194, p < .001$). The narrow confidence interval shown in the forest plot suggests this is a precise and reliable predictor.

- Average Review Score: Surprisingly, the model revealed a significant negative relationship between average review scores and listens ($\beta = -12.836, p = .009$). As visualized in the forest plot, the $95\%$ confidence interval for this estimate remains entirely below zero, confirming a statistically significant downward trend.

- Model Diagnostics: Residual analysis plots indicate that the assumptions of linearity and homoscedasticity were generally met, although some outliers are visible at higher listen counts.

In real-world terms, the analysis suggests that for a podcast's popularity, engagement volume matters far more than the average rating.

- Popularity Breeds Reviews: As expected, podcasts with a higher total number of reviews tend to have more daily listens. For every additional review received, we expect an average increase of approximately $0.19$ daily listens.

- The "Rating Paradox": Counter-intuitively, podcasts with higher average scores actually tend to have slightly fewer daily listens. This suggests that a perfect 5-star rating is often a sign of a small, niche audience. Larger, more popular podcasts likely attract a more diverse range of listeners, which naturally pulls the average score down even as the total number of listeners grows.

- Other Factors: Because the $R^{2}$ value is very low, it is important to note that reviews are not the main driver of podcast success. The vast majority of a podcast's popularity is likely determined by factors not included in this specific test, such as the genre or the platform used.

### Genre Analysis and Subscriber Interaction

This section contrasts listenership across podcast categories and investigates how genre moderates the relationship between subscribers and listens.

1. Contrasting Listens Across Genres
- To evaluate if daily.listens differ by genre, the following methods were used:

- Visual Contrast: A boxplot was generated to compare medians, spreads, and outliers across genres.

- Inferential Analysis: A one-way ANOVA tested for significant differences in mean listens.

- Post-hoc Testing: Tukey’s HSD was applied to identify specific genre pairs with statistically significant differences, following standard reporting conventions.

2. Interaction Effect: Subscribers and Genre
- The final analysis examined if the effect of subscribers on listens varies by genre.

- Interaction Modeling: A linear model using the subscribers * genre operator was fitted to test for varying slopes.

- Slope Visualization: An interaction plot with regression lines per genre was created. Differing gradients visually demonstrate whether certain genres gain more listens per subscriber than others.

- Validation: Residual plots were generated to ensure the interaction model satisfied assumptions of linearity and homoscedasticity.

This approach demonstrates Technical Capability  by moving beyond simple correlations to complex interactions. By documenting these steps, the report allows another analyst to verify the logic and "easily modify the code" for updated data.

```{r}
# Visualizing listens across genres
ggplot(podcast_clean, aes(x = genre, y = daily.listens, fill = genre)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "Distribution of Daily Listens by Genre",
    x = "Podcast Genre",
    y = "Daily Listens"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates labels for readability

# Analysis: One-way ANOVA to compare means across genres
genre_model <- aov(daily.listens ~ genre, data = podcast_clean)

# View the results
summary(genre_model)

# If significant, use Tukey's HSD to see which specific genres differ
TukeyHSD(genre_model)

# Interaction model: Listens predicted by Subscribers, Genre, and their interaction
interaction_model <- lm(daily.listens ~ subscribers * genre, data = podcast_clean)

# Summary of the interaction
summary(interaction_model)

# Visualization of the interaction effect
ggplot(podcast_clean, aes(x = subscribers, y = daily.listens, color = genre)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) + # Regression lines per genre
  labs(
    title = "Interaction Effect: Subscribers vs Listens by Genre",
    x = "Number of Subscribers",
    y = "Daily Listens"
  ) +
  theme_light()
```
1. Contrast Across Genres (ANOVA & Boxplot)

A one-way ANOVA was conducted to evaluate the effect of genre on daily listenership. The analysis revealed highly significant differences between categories. Post-hoc Tukey HSD testing confirmed that all genre pairs (Comedy, Documentary, and News & Culture) differ significantly from one another ($p < .001$ for all comparisons). As visualized in the boxplot, Comedy maintains the highest median daily listens, followed by News & Culture, with Documentary exhibiting the lowest average listenership.

The data shows that a podcast's genre has a major impact on how many people listen to it3. Comedy is clearly the most popular category in this dataset, while Documentaries tend to attract the smallest daily audiences. These differences are not due to random chance; there are distinct "tiers" of popularity based on the type of content being produced.

2. Effect of Subscribers by Genre (Interaction Model)

A multiple linear regression with an interaction term was used to determine if the effect of subscribers on listens varies by genre4. The model is highly robust, explaining $61.25\%$ of the variance ($Adjusted R^{2} = 0.6125$).

- Main Effect: Across all genres, there is a significant positive relationship between subscribers and daily listens ($\beta = 0.326, p < .001$).

- Interaction Effect: The interaction between subscribers and the Documentary genre is significant ($\beta = -0.079, p = .0114$), as shown by the flatter green slope in the interaction plot. This indicates that the "benefit" of each additional subscriber is significantly lower for Documentaries compared to the Comedy baseline.

While having more subscribers generally leads to more daily listens, the "power" of a subscriber varies by genre. In Comedy and News & Culture, gaining new subscribers provides a strong, consistent boost to daily listens (steeper slopes). However, for Documentaries, each new subscriber results in a much smaller increase in actual daily activity (flatter slope). This suggests that Documentary subscribers may be less frequent or habitual listeners than fans of Comedy or News.

### Subscriber Impact Moderated by Genre

To address the final requirement of Question 2, a multiple linear regression with an interaction term (subscribers * genre) was utilized. This determines if the relationship between subscribers and listens varies significantly by podcast category.

- Interaction Modeling: By including the interaction operator, the model allows for different slopes per genre, testing if the "return" on each subscriber is moderated by the content type.

- Visual Analysis: An interaction plot was created using geom_smooth(method = "lm") to display these varying slopes.

- Diagnostic Validation: Residual analysis was performed using a $2 \times 2$ grid (Residuals vs Fitted, Normal Q-Q, Scale-Location, and Residuals vs Leverage) to ensure model integrity by verifying linearity and homoscedasticity.

```{r}
# Analysis: Does the relationship between subscribers and listens depend on the genre?
model_interaction <- lm(daily.listens ~ subscribers * genre, data = podcast_clean)

# Display the results professionally
summary(model_interaction)

# Visualization: Interaction Plot
ggplot(podcast_clean, aes(x = subscribers, y = daily.listens, color = genre)) +
  geom_point(alpha = 0.2) + # Faded points to show density
  geom_smooth(method = "lm", se = FALSE, size = 1.2) + # Linear slopes per genre
  labs(
    title = "Interaction Effect: Subscribers vs. Listens by Genre",
    subtitle = "Varying slopes indicate that subscriber impact differs by category",
    x = "Number of Subscribers",
    y = "Daily Listens",
    color = "Genre"
  ) +
  theme_minimal()

# Set up a 2x2 grid for diagnostic plots
par(mfrow = c(2, 2))

# Generate diagnostic plots for the interaction model
plot(model_interaction)

# Reset the plotting layout to default (1x1)
par(mfrow = c(1, 1))
```
#### Interaction Model Results (NHST & Estimation)

The model is highly robust, explaining $61.25\%$ of the variance in daily listens ($Adjusted R^{2} = 0.6125$).

- Main Effect of Subscribers: Across the baseline category, there is a strong, highly significant positive relationship between subscribers and daily listens ($\beta = 0.326, p < 2 \times 10^{-16}$).

- Interaction Effect: The interaction between subscribers and the Documentary genre is statistically significant and negative ($\beta = -0.079, p = .0114$). This indicates that the "return" on each additional subscriber is significantly lower for Documentaries compared to the Comedy baseline.

- Interaction Effect: The interaction for News & Culture was positive but did not reach the threshold for statistical significance at the $5\%$ level ($\beta = 0.052, p = .0842$), suggesting its subscriber-to-listen ratio is more comparable to the baseline.

#### Visual Interpretation of Interaction Plot

The Interaction Plot confirms these findings by displaying varying slopes for each genre:

- High Engagement: Comedy (red) and News & Culture (blue) exhibit steeper gradients, indicating that subscriber growth in these categories translates effectively into higher daily listens.

- Lower Engagement Impact: The Documentary genre (green) displays a notably flatter slope. Visually, this confirms the significant interaction term; gaining 500 subscribers in the Documentary genre provides a much smaller boost to daily listenership than it does for the other categories.

#### Model Diagnostic Validation

The Residual Analysis grid confirms that the interaction model is statistically sound and meets the primary assumptions of linear regression:

- Linearity (Residuals vs Fitted): The red line is relatively horizontal, indicating that the relationship between predictors and the outcome is correctly modeled as linear.

- Normality (Normal Q-Q): The residuals follow the dashed diagonal line closely, confirming that the error terms are normally distributed.

- Homoscedasticity (Scale-Location): The consistent spread of points suggests that the variance of the residuals is constant across the range of fitted values.

- Influential Observations (Residuals vs Leverage): All data points fall within the limits of Cook's distance, indicating that no single podcast is exerting an undue influence on the model’s coefficients.


### Overall Summary

The analysis reveals that podcast popularity is driven primarily by engagement volume and content category rather than high average ratings. While total review volume and subscriber counts are strong positive indicators of daily listens, average review scores surprisingly show a negative relationship ($\beta = -12.836, p = .009$), suggesting that broad-reach podcasts naturally accumulate more diverse (and lower) ratings than niche programs.

Key Analytical Findings:

- Engagement Over Rating: A multiple linear regression ($R^2 = 0.029$) confirmed that every additional review predicts a significant increase in listens, while a higher average score actually corresponds with a decrease in traffic.

- Genre-Based Disparity: Significant differences in listenership exist across genres ($p < .001$), with Comedy maintaining the highest median daily listens, followed by News & Culture and Documentary.

- Moderated Subscriber Effect: An interaction model ($Adjusted R^{2} = 0.6125$) proves that the "return" on subscribers varies by genre. Gaining subscribers provides a strong, consistent boost for Comedy and News & Culture, but has a significantly lower impact on Documentaries ($\beta = -0.079, p = .0114$), indicating a more passive audience for that category.

- Technical Validation: All findings were validated through residual analysis; diagnostic plots confirmed the model's integrity by meeting the assumptions of linearity, normality, and homoscedasticity without influential outliers.

Conclusion for Business Strategy:

For commercial growth, creators should focus on total audience acquisition and review volume rather than maintaining a perfect 5-star rating. Furthermore, resource allocation for subscriber growth should be prioritized for highly responsive genres like Comedy and News, as Documentary audiences show lower incremental engagement per subscriber.


# References

- (Geeks for Geeks. (2025) Referencing. [online] Available from: https://www.geeksforgeeks.org/r-language/how-to-do-a-tukey-hsd-test-with-the-anova-command-car-package-in-r/ (Accessed 23 July 2025)

- (RPubs by RStudio. (2023) Referencing. [online] Available from: https://rpubs.com/sharmaar3/Residual_Plots)



